Biogeography of crop progenitors and wild plant resources in West Asia, 21–8.3 ka

### 2025-01-04 analysis/paper.qmd
Optimise model code

---
title: Biogeography of crop progenitors and wild plant resources during the transition to agriculture in West Asia, 21–8.3 ka
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This paper presents the first comprehensive reconstructions of the 
  palaeodistributions of [X] plant species known to have been gathered or 
  cultivated by early agricultural societies in Southwest Asia, including the 
  progenitors of the first crops. We used machine learning to train an 
  ecological niche model (ENM) of each species based on its present-day 
  distribution in relation to climate and environmental variables. Predictions 
  of the potential ranges of these species at key stages of the 
  Pleistocene–Holocene transition could then be derived from these models using 
  hindcast data from palaeoclimate simulations.
  Species ranges were on average [X%] [larger|smaller] in the Early Holocene 
  compared to today, indicating [...]. The modelled ranges predict the
  observed occurrence of these species on archaeological sites with [low|medium|high]
  accuracy. The regional ubiquity of species in the archaeological record is [not] 
  correlated with the predicted size of its range and the diversity of
  archaeobotanical assemblages is [not] correlated with the predicted diversity
  of its environs. This indicates that trends in taxonomic composition of
  the archaeobotanical record is [not] likely to be influenced by environmental 
  change and species turnover, not just, as is often assumed, human economic 
  choices.
format: 
  elsevier-pdf:
    keep-tex: true
    cite-method: citeproc
    journal:
      name: Open Quaternary
      formatting: review
bibliography: references.bib
execute: 
  echo: false
  eval: false
---

* [ ] Finish analysis:
  - [ ] Add terrain, soil
  - [ ] Find a resolution that's workable
  - [ ] Do thresholding?
- [ ] First draft:
  - [ ] Introduction
  - [ ] Background
    - [ ] Biogeography
    - [ ] ENM
  - [x] Methods & Materials
    - [x] Occurrence data
    - [x] Predictor data
    - [x] Random Forest model
  - [ ] Results
    - [ ] Model assessment
    - [ ] Hindcasting
  - [ ] Discussion
    - [ ] General trends
    - [ ] Case studies
  - [ ] Conclusion
- [ ] Figures
  - [ ] tbl-predictors
  - [ ] tbl-climate-periods
* [ ] Appendix with all hindcast predictions
- [ ] References & copyediting
- [ ] Final proofread

```{r dependencies, eval=TRUE, message=FALSE}
library("cowplot")
library("dplyr")
library("dotenv")
library("english")
library("forcats")
library("fs")
library("furrr")
library("ggplot2")
library("ggrepel")
library("ggspatial")
library("gt")
library("here")
library("khroma")
library("progressr")
library("purrr")
library("ranger")
library("readr")
library("recipes")
library("rgbif")
library("scales")
library("sf")
library("stars")
library("stringr")
library("tibble")
library("tidymodels")
library("tidyr")

@@

```{r setup-whitebox}
```

```{r define-regions}
latlong <- 4326
utm37n <- 32637

@@

```{r define-periods, eval=TRUE}
archaeo_periods <- tribble(
  ~period,             ~agriculture,                          ~start_bp, ~end_bp,
  "Late Epipal.",      "Foraging",                            15000,     11700,
  "PPNA",              "Pre-domestication cultivation",       11700,     10700,
  "EPPNB",             "Cultivation of domesticated species", 10700,     10200,
  "MPPNB",             "Cultivation of domesticated species", 10200,     9500,
  "LPPNB/C",           "Agriculture",                         9500,      8500,
  "Pottery Neolithic", "Agriculture",                         8500,      6500,
  "Chalcolithic",      "Agriculture",                         6500,      5000,
)

@@

climate_periods <- fct_inorder(c(
  lgm = "LGM (21 ka)",
  hs1 = "HS1 (17.0-14.7 ka)",
  ba  = "BA (14.7-12.9 ka)",
  yds = "YDS (12.9-11.7 ka)",
  eh  = "EH (11.7-8.3 ka)",
  cur = "current"
))
```

@@

```{r read-basemap}
# Turn off s2 while we deal with Natural Earth's dodgy geometries
use_s2 <- sf_use_s2(FALSE)

@@

@@

```{r data-flora}
# Flora present at archaeological sites
archaeo_flora <- read_tsv(raw_data("swasia_neolithic_flora.tsv")) |>
  rename(taxon_group = "taxon", taxon = "taxon_detail") |>
  # Assign to archaeological periods
  # TODO: refactor as function
  # TODO: assign climate periods too
  mutate(
    age_mid = age_end + ((age_start - age_end) / 2),
    period = cut(-age_mid, 
                 breaks = -c(archaeo_periods$start_bp, 0), 
                 labels = archaeo_periods$period,
                 ordered_result = TRUE)
  ) ->
  archaeo_flora

@@

n_site_neolithic <- archaeo_flora |> 
  filter(period %in% neolithic) |> 
  distinct(site_name) |> 
  nrow()

# Flora to be modelled
archaeo_flora |>
  filter(period %in% neolithic) |>
  group_by(taxon) |>
  summarise(
    n_present = length(unique(phase_code)),
    p_present = n_present / n_assemb_neolithic,
    .groups = "drop"
  ) |>
  filter(n_present > 2) |>
  # Exclude taxa not identified to species
  filter(!str_ends(taxon, "sp.")) ->
  flora
```

@@

```{r data-flora-taxonomy}
# - Normalise species names and match to the GBIF Backbone Taxonomy
# - Recode domesticates as their wild progenitors.
# - Add other wild progenitors of wheats
flora <- flora |>
  mutate(
    taxon = case_match(taxon,
                       "Vicia narbonense" ~ "Vicia narbonensis", # N=3
                       "Triticum aestivocompactum" ~ "Triticum aestivum compactum", # N=3
                       "Heliotropium persicum" ~ "Moltkia angustifolia", # N=2
                       "Lens orientalis" ~ "Vicia orientalis", # N=2
                       "Scirpus tabernaemontani" ~ "Schoenoplectus tabernaemontani", # N=2
                       "Stipa capensis" ~ "Stipa dregeana", # N=2
                       .default = taxon),
    taxon_original = taxon,
    taxon = map_chr(taxon, gbif_accepted_name),
    taxon = case_match(
      taxon,
      "Cicer arietinum" ~ "Cicer reticulatum",
      "Hordeum vulgare" ~ "Hordeum spontaneum",
      "Vicia lens" ~ "Vicia orientalis",
      "Linum usitatissimum" ~ "Linum bienne",
      "Triticum aestivum" ~ "Triticum turgidum dicoccum",
      "Triticum dicoccoides" ~ "Triticum turgidum dicoccum",
      "Triticum turgidum durum" ~ "Triticum turgidum dicoccum",
      "Triticum monococcum" ~ "Triticum monococcum aegilopoides",
      .default = taxon
    )
  ) |>
  bind_rows(tibble(taxon = c("Triticum urartu", "Aegilops speltoides", 
                             "Aegilops tauschii")))
```

Taxonomic names were resolves to the canonical form specified in the GBIF Backbone Taxonomy [@GBIFSecretariat2023].
<!-- TODO: still relevent? aren't we modelling species? -->So for example occurrences for "Bolboschoenus" included all species and subspecies, including specimens described as *Bolboschoenus sp.*, *Bolboschoenus maritimus* and *Scirpus maritimus*.
Domestic species meeting our inclusion criteria were substituted with their wild progenitor(s), where known.

@@

```{r data-occ-count}
flora |>
  group_by(taxon) |>
  summarise(
    taxon_original = list(c(taxon_original)),
    n_present = max(n_present), # TODO: be more precise?
    p_present = max(p_present)
    ) |>
  mutate(n_occ = map_dbl(taxon, \(x) occ_count(scientificName = x, 
                                               hasCoordinate = TRUE, 
                                               hasGeospatialIssue = FALSE,
                                               geometry = bbox_wkt(w_asia)))) ->
  flora
```

```{r tbl-occ-count}
#| tbl-cap: "Species modelled in this study."
# TODO: add to caption:
# - with presence in Neolithic sites in West Asia, out of the X sites included in Y
# - with number of reliably-georeferenced occurrences in West Asia recorded in GBIF
# TODO: add count of species included?
flora |>
  mutate(
    taxon_detail = map_chr(taxon_original, paste, collapse = ", "),
    taxon_detail = if_else(taxon == taxon_detail, NA_character_, taxon_detail),
    #include = n_occ > 50
  ) |>
  select(-taxon_original) |>
  # group_by(include) |>
  filter(n_occ > 50) |>
  arrange(-n_present) |>
  gt() |>
  # TODO: move synonyms to footnotes?
  cols_merge(
    c(taxon, taxon_detail),
    rows = !is.na(taxon_detail),
    pattern = "{1}<br><small>(incl. {2})</small>"
  ) |>
  fmt_percent(p_present, decimals = 0) |>
  cols_merge_n_pct(n_present, p_present) |>
  tab_style(cell_text(style = "italic"), cells_body(taxon))
```

```{r filter-flora}
flora <- filter(flora, n_occ > 50)
# Why are there 47k occurrences of Avena sterilis??
flora <- filter(flora, taxon != "Avena sterilis")
```

Occurrence data was obtained from the Global Biodiversity Information Facility (GBIF) using via its application programming interface and the R package rgbif [@ChamberlainBoettiger2017; @rgbif].
Although ENMs have reasonable predictive power even with small training samples [@StockwellPeterson2002; @HernandezEtAl2006; @WiszEtAl2008], we excluded X<!-- TODO --> taxa with less than fifty recorded occurrences in our study region.
We also excluded one taxon (*Avena sterilis*) with over 47,000 occurrences, as this would have been computationally prohibitive and we were uncertain what account for such a disproportionately high number of records.

```{r fetch-occ}
# TODO: Refactor
gbif_data <- function(scientificName, ...) {
  gbif_data <- rgbif::occ_data(scientificName = scientificName, ...)
  gbif_data$data
}

flora |>
  mutate(
    occ = map(taxon, gbif_data, 
              hasCoordinate = TRUE, hasGeospatialIssue = FALSE, 
              geometry = w_asia, limit = 100000,
              .progress = TRUE)
  ) ->
  flora
```

```{r tidy-occ}
# TODO: further tidying. Ideas:
# - https://data-blog.gbif.org/post/gbif-filtering-guide/
# - https://cran.r-project.org/web/packages/CoordinateCleaner/vignettes/Cleaning_GBIF_data_with_CoordinateCleaner.html
# - https://www.r-bloggers.com/2021/03/downloading-and-cleaning-gbif-data-with-r/
flora |>
  mutate(
    occ = map(occ, gbif_drop_imprecise_coords, threshold = 5000),
    occ = map(occ, gbif_drop_fossils),
    occ = map(occ, gbif_drop_duplicate_coords),
    occ = map(occ, gbif_standardise_names),
    occ = map(occ, \(x) {
      select(x, gbif_key = key, scientific_name, genus, species, longitude, 
             latitude)
    })
  ) ->
  flora
```

```{r fig-occ-map}
flora |>
  unnest(occ) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  ggplot() + 
  facet_wrap(vars(taxon)) +
  layer_spatial(ne_land, fill = "#ffffff") +
  geom_sf(shape = ".") +
  coord_sf(xlim = w_asia[c("xmin", "xmax")], 
           ylim = w_asia[c("ymin", "ymax")],
           crs = utm37n,
           default_crs = latlong)
```

<!-- TODO: occurrence tidying -->

```{r generate-background}
# TODO: maybe defer this to just before the model is actually fit ?
#   (and discard the background sample)
# using a custom recipe step even?
n_background <- 10000

flora <- flora |>
  mutate(
    occ = map(occ, mutate, present = TRUE, weight = 1),
    bg_data = map(occ, function(x, region, n) { 
      background_sample(
        region = region, 
        N = n,
        coord_x = "longitude", coord_y = "latitude",
        present = FALSE,
        weight = (1 / n) * nrow(x) # weight presences and background equally
      )
    }, region = w_asia_land, n = n_background),
    occ = map2(occ, bg_data, bind_rows)
  ) |>
  select(-bg_data)
```

<!-- TODO: fix informal tone -->
Occurrence data only tells us where a species is present;
there is rarely definitive information on where the species is *not* found.
We therefore need to generate random background points or "pseudo-absences" to feed to the model.
There are several ways to do this. 
We follow the advice of @BarbetMassinEtAl2012 for regression-based species distribution models and use a large (:10000) random sample of points, weighted equally against the presences in the regression.
@ValaviEtAl2022 also recommend using a very large background sample for random forest models.

@@

climate <- read_stars(
  derived_data(
    "paleoclim", 
    paste0("paleoclim_", names(climate_periods), "_2_5m_w_asia.tif")
  ),
  along = NA
)

names(climate) <- names(climate_periods)

@@

# TODO: derived terrain products are two big to commit
# TODO: does TWI actually make sense at this resolution? Could start with (much)
# higher res, then upscale?
terrain <- read_stars(derived_data("dem", c("srtm_15plus_w_asia.tif", 
                                            "srtm_15plus_w_asia_slope.tif",
                                            "srtm_15plus_w_asia_aspect.tif",
                                            "srtm_15plus_w_asia_wetness.tif")))
names(terrain) <- c("elevation", "slope", "aspect", "wetness")
```

@@

```{r data-environment}
# For time-invariant predictors, collapse attributes into dimension and 
# duplicate to create a dummy period dimension
# TODO: maybe need to downsample first?
terrain <- st_redimension(terrain, name = "variable")
terrain <- do.call(c, rep(list(terrain), length(climate_periods)))
names(terrain) <- names(climate_periods)
terrain <- st_redimension(terrain, name = "period") # TODO: why *so* slow?

soil <- st_redimension(soil, name = "variable")
soil <- do.call(c, rep(list(soil), length(climate_periods)))
names(soil) <- names(climate_periods)
soil <- st_redimension(soil, name = "period")

# Resample to common grid
terrain <- st_warp(terrain, climate, use_gdal = TRUE, method = "cubicspline")
soil <- st_warp(soil, climate, use_gdal = TRUE, method = "cubicspline")

# TODO: combine and split variable
# TODO: fill NAs?
```

@@

```{r extract-predictor-data}
flora |> 
  mutate(
    occ = map(occ, \(x) cbind(x, st_extract(climate$cur, gbif_to_sf(x)))),
    # TODO: where is the extra geometry column coming from?
    occ = map(occ, \(x) as_tibble(select(x, -geometry))),
    # TODO: can this happen earlier?
    occ = map(occ, \(x) mutate(x, present = factor(present, c(FALSE, TRUE))))
  ) ->
  flora
```

@@

```{r niche-data}
# Split training, test (for model assessment) and cross-validation sets (for
# tuning)
flora <- flora |>
  mutate(
    occ = map(occ, \(x) mutate(x, weight = importance_weights(weight))),
    occ = map(occ, \(x) drop_na(x, starts_with("bio"))), # TODO: where do these come from?
    occ_split = map(occ, \(x) initial_split(x, strata = present)),
    # occ_folds = map(occ_split, \(x) vfold_cv(training(x), v = 10, strata = present))
  )

# Data preprocessing recipe
# Following recommended preprocessing steps for RF classification from:
#   https://www.tmwr.org/pre-proc-table.html
niche_recipe <- recipe(training(flora$occ_split[[1]])) |>
  update_role(gbif_key:latitude, new_role = "ID") |>
  update_role(present, new_role = "outcome") |>
  update_role(starts_with("bio"), new_role = "predictor") |>
  step_nzv(all_predictors())
```

@@

# TODO: Or... tune parameters? But this is very compute-intensive.
# niche_model <- rand_forest(
#   mode = "classification", 
#   mtry = tune(),
#   trees = tune(),
#   min_n = tune()
# ) |>
#   set_engine("ranger", probability = TRUE)

niche_workflow <- workflow(niche_recipe, niche_model)

# Tune model (separately for each taxon)
# TODO: questioning - would be nice but is extremely slow
# flora |>
#   slice_sample(n = 10) |>
#   mutate(
#     niche_tuning_grid = map(occ_split, function(data, recipe) {
#       grid_regular(
#         finalize(mtry(), bake(prep(recipe), training(data), all_predictors())),
#         trees(),
#         min_n(),
#         levels = 5
#       )
#     }, recipe = niche_recipe, .progress = TRUE),
#     niche_tuning = map2(niche_tuning_grid, occ_folds, function(grid, folds, wflow) {
#       tune_grid(wflow, folds, grid = grid)
#     }, wflow = niche_workflow, .progress = TRUE),
#     niche_hparams = map(niche_tuning, select_best, metric = "roc_auc"),
#     niche_model = map(niche_hparams, function(params, wflow) {
#       finalize_workflow(wflow, params)
#     }, wflow = niche_workflow)
#   )
```

@@

```{r fig-pred-observed-eh}

```

>>>>

---
title: Biogeography of crop progenitors and wild plant resources in West Asia, 21–8.3 ka
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This paper presents the first comprehensive reconstructions of the 
  palaeodistributions of 56 plant species found regularly in association with
  early agricultural archaeological sites in West Asia, including the 
  progenitors of the first crops. We used machine learning to train an 
  ecological niche model of each species based on its present-day distribution 
  in relation to climate and environmental variables. Predictions of the 
  potential ranges of these species at key stages of the Pleistocene–Holocene 
  transition could then be derived from these models using hindcast data from 
  palaeoclimate simulations.
  Species ranges are predicted to have been on average [X%] [larger|smaller] in 
  the Early Holocene compared to present conditions, indicating [...]. 
  The modelled ranges predict the observed occurrence of these species on 
  archaeological sites with [low|medium|high] accuracy. The regional ubiquity of 
  species in the archaeological record is [not] correlated with the predicted 
  size of its range and the diversity of archaeobotanical assemblages is [not] 
  correlated with the predicted diversity of its environs. This indicates that 
  trends in taxonomic composition of the archaeobotanical record is [not] likely 
  to have been influenced by environmental change and species turnover, in
  addition to human economic choices.
format: 
  elsevier-pdf:
    keep-tex: true
    cite-method: citeproc
    journal:
      name: Open Quaternary
      formatting: review
bibliography: references.bib
execute: 
  echo: false
---

* [x] Finish modelling
- [ ] First draft:
  - [ ] Introduction
  - [ ] Background
    - [ ] Biogeography
    - [x] ENM
  - [x] Methods & Materials
  - [ ] Results
    - [ ] Model assessment
    - [ ] Hindcasting
  - [ ] Discussion
    - [ ] General trends
    - [ ] Case studies
  - [ ] Conclusion
- [ ] Figures
  - [ ] tbl-occ-count
  - [ ] tbl-predictors
  - [ ] tbl-climate-periods
- [ ] Clean up package code
- [ ] Appendix with all hindcast predictions
- [ ] References & copyediting
- [ ] Final proofread

```{r dependencies, message=FALSE}
library("cowplot")
library("dplyr")
library("dotenv")
library("english")
library("forcats")
library("fs")
library("furrr")
library("ggplot2")
library("ggrepel")
library("ggspatial")
library("gt")
library("here")
library("khroma")
library("progressr")
library("purrr")
library("ranger")
library("readr")
library("recipes")
library("rgbif")
library("scales")
library("sf")
library("stars")
library("stringr")
library("tibble")
library("tidymodels")
library("tidyr")

@@

```{r define-regions}
latlong <- 4326
w_asia_albers <- "+proj=aea +lat_1=22.5 +lat_2=42.5 +lon_0=40"

@@

```{r define-periods}
archaeo_periods <- tribble(
  ~period,             ~agriculture,                          ~start_bp, ~end_bp,
  "Late Epipal.",      "Foraging",                            15000,     11700,
  "PPNA",              "Pre-domestication cultivation",       11700,     10700,
  "EPPNB",             "Cultivation of domesticated species", 10700,     10200,
  "MPPNB",             "Cultivation of domesticated species", 10200,     9500,
  "LPPNB/C",           "Agriculture",                         9500,      8500,
  "Pottery Neolithic", "Agriculture",                         8500,      6500,
  "Chalcolithic",      "Agriculture",                         6500,      5000,
)

@@

climate_periods <- tibble(
  period = c("Last Glacial Maximum", "Heinrich Stadial 1", "Bølling-Allerød",
             "Younger Dryas", "Early Holocene", "Current"),
  code = c("lgm", "hs1", "ba", "yds", "eh", "cur"),
  label = fct_inorder(c("LGM (21 ka)", "HS1 (17.0-14.7 ka)", "BA (14.7-12.9 ka)", 
            "YDS (12.9-11.7 ka)", "EH (11.7-8.3 ka)", "current")),
  start_bp = c(21000, 17000, 14700, 12900, 11700, NA),
  end_bp = c(17000, 14700, 12900, 11700, 8300, NA)
)
```

```{r define-parameters}
# Minimum number of GBIF occurrences for modelling
min_n_occ <- 50

# Number of background samples per taxon
n_background <- 10000

# Raster resolution for prediction (meters)
pred_res <- 5000
```

@@

```{r data-basemap, message=FALSE}
# Turn off s2 while we deal with Natural Earth's dodgy geometries
use_s2 <- sf_use_s2(FALSE)

@@

```{r fig-region}
#| fig-cap: Study region
ggplot() + 
  annotation_spatial(ne_land, fill = "#ffffff") +
  layer_spatial(w_asia, fill = NA) +
  coord_sf(crs = w_asia_albers)
```

<!-- TODO define study region and period -->

@@

```{r data-archaeo-flora}
# Flora present at archaeological sites
archaeo_flora <- read_tsv(
  raw_data("swasia_neolithic_flora.tsv"),
  col_types = cols(
    source = col_character(),
    source_id = col_character(),
    source_site_name = col_character(),
    site_name = col_character(),
    latitude = col_double(),
    longitude = col_double(),
    phase = col_character(),
    phase_description = col_character(),
    phase_code = col_character(),
    age_start = col_integer(),
    age_end = col_integer(),
    taxon_source = col_character(),
    n = col_double(),
    prop = col_double(),
    reference = col_character(),
    taxon_detail = col_character(),
    taxon = col_character(),
    genus = col_character(),
    family = col_character(),
    category = col_character(),
    founder_crop = col_character(),
    edibility = col_character(),
    grass_type = col_character(),
    legume_type = col_character()
  )
) |>
  rename(taxon_group = "taxon", taxon = "taxon_detail") |>
  # Assign to archaeological periods
  mutate(
    age_mid = age_end + ((age_start - age_end) / 2),
    period = cut(
      -age_mid, 
      breaks = -c(archaeo_periods$start_bp, 0), 
      labels = archaeo_periods$period,
      ordered_result = TRUE
    ),
    climate_period = cut(
      -age_mid, 
      breaks = -c(climate_periods[climate_periods$code != "cur",]$start_bp, 0), 
      labels = climate_periods[climate_periods$code != "cur",]$code,
      ordered_result = TRUE
    )
  ) ->
  archaeo_flora

@@

n_site_neolithic <- archaeo_flora |> 
  filter(period %in% neolithic) |> 
  distinct(site_name) |> 
  nrow()
```

@@

```{r data-flora}
if (!file_exists(derived_data("flora.tsv"))) {
  message("Regenerating flora...")
  # Subset of flora to be modelled
  archaeo_flora |>
    filter(period %in% neolithic) |>
    group_by(taxon) |>
    summarise(
      n_present = length(unique(phase_code)),
      p_present = n_present / n_assemb_neolithic,
      .groups = "drop"
    ) |>
    filter(n_present > 2) |>
    # Exclude taxa not identified to species
    filter(!str_ends(taxon, "sp.")) ->
    flora

  # Standardise taxonomy
  flora <- flora |>
    # Normalise species names and match to the GBIF Backbone Taxonomy
    mutate(
      taxon = case_match(taxon,
                         "Vicia narbonense" ~ "Vicia narbonensis", # N=3
                         "Triticum aestivocompactum" ~ "Triticum aestivum compactum", # N=3
                         "Heliotropium persicum" ~ "Moltkia angustifolia", # N=2
                         "Lens orientalis" ~ "Vicia orientalis", # N=2
                         "Scirpus tabernaemontani" ~ "Schoenoplectus tabernaemontani", # N=2
                         "Stipa capensis" ~ "Stipa dregeana", # N=2
                         .default = taxon),
      taxon_original = taxon,
      taxon = map_chr(taxon, gbif_accepted_name)
    ) |>
    # Recode domesticates as their wild progenitors.
    mutate(
      taxon = case_match(
        taxon,
        "Cicer arietinum" ~ "Cicer reticulatum",
        "Hordeum vulgare" ~ "Hordeum spontaneum",
        "Vicia lens" ~ "Vicia orientalis",
        "Linum usitatissimum" ~ "Linum bienne",
        "Triticum aestivum" ~ "Triticum turgidum dicoccum",
        "Triticum dicoccoides" ~ "Triticum turgidum dicoccum",
        "Triticum turgidum durum" ~ "Triticum turgidum dicoccum",
        "Triticum monococcum" ~ "Triticum monococcum aegilopoides",
        .default = taxon
      )
    ) |>
    # Add wild progenitors of wheats
    bind_rows(
      tibble(
        taxon = c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii"),
        taxon_original = c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii")
      )
    ) |>
    # Deduplicate by canonical taxon
    summarise(
      taxon_detail = paste0("*", taxon_original[taxon_original != taxon], "*", collapse = ", "),
      taxon_detail = na_if(taxon_detail, "**"),
      n_present = max(n_present),
      p_present = max(p_present),
      .by = taxon
    )

  write_tsv(flora, derived_data("flora.tsv"))
}

flora <- read_tsv(
  derived_data("flora.tsv"),
  col_types = cols(
    taxon = col_character(),
    taxon_detail = col_character(),
    n_present = col_double(),
    p_present = col_double()
  )
)
```

Taxonomic names were resolves to the canonical form specified in the GBIF Backbone Taxonomy [@GBIFSecretariat2023].
So for example occurrences for *Bolboschoenus maritimus* also include those recorded under the older nomenclature *Scirpus maritimus* (see @tbl-occ-count).
Domestic species meeting our inclusion criteria were substituted with their wild progenitor(s), where known.

@@

```{r data-occ-count}
if (!file_exists(derived_data("flora_occ_count.tsv"))) {
  message("Reacquiring GBIF occurrence counts...")
  flora <- flora |>
    mutate(
      n_occ = map_dbl(taxon, \(x) occ_count(scientificName = x, 
                                            hasCoordinate = TRUE, 
                                            hasGeospatialIssue = FALSE,
                                            geometry = bbox_wkt(w_asia)))
    )
  write_tsv(flora, derived_data("flora_occ_count.tsv"))
}
flora <- read_tsv(derived_data("flora_occ_count.tsv"))
```

```{r tbl-occ-count}
#| tbl-cap: "Recorded occurrence of flora considered in this study in Neolithic and contemporary West Asia"
flora |>
  mutate(
    exclude = (n_occ < min_n_occ) | (taxon == "Avena sterilis"),
    label = if_else(
      !is.na(taxon_detail),
      glue::glue("*{taxon}*\n\n<small>incl. {taxon_detail}</small>"),
      paste0("*", taxon, "*")
    )
  ) |>
  select(label, n_present, p_present, n_occ, exclude) |>
  arrange(-n_present) |>
  gt() |>
  cols_label(
    label = "Taxon",
    n_present = "Neolithic",
    n_occ = "Present"
  ) |>
  tab_spanner("Occurrences", columns = starts_with("n_")) |>
  cols_merge_n_pct(n_present, p_present) |>
  cols_hide(exclude) |>
  tab_footnote(
    "Excluded from modelling due to sample size", 
    cells_body(columns = c(n_occ), rows = exclude),
    placement = "right"
  ) |>
  cols_align("left", c(label)) |>
  fmt_percent(p_present, decimals = 0) |>
  fmt_markdown(label) |>
  sub_missing() |>
  tab_options(footnotes.marks = "*")
```

Georeferenced occurrence data was obtained from the Global Biodiversity Information Facility (GBIF) using via its application programming interface and the R package 'rgbif' [@ChamberlainBoettiger2017; @rgbif].
GBIF occurrences marked has having imprecise or duplicate coordinates were excluded from the training dataset, as were fossil records.
Although niche models have reasonable predictive power even with small training samples [@StockwellPeterson2002; @HernandezEtAl2006; @WiszEtAl2008], we excluded `r english(nrow(filter(flora, n_occ > min_n_occ)))` taxa with less than `r min_n_occ` occurrences in West Asia, following recommendations for niche models generally and Random Forest-based models specifically [@StockwellPeterson2002; @LuanEtAl2020].
We also excluded one taxon (*Avena sterilis*) with over 50,000 occurrences, as this would have been computationally prohibitive and we were uncertain what account for such a disproportionately high number of records.

```{r filter-flora}
flora <- filter(flora, n_occ > min_n_occ)
# Why are there 50k occurrences of Avena sterilis??
flora <- filter(flora, taxon != "Avena sterilis")
```

```{r fetch-occ}
# TODO: further tidying. Ideas:
# - https://data-blog.gbif.org/post/gbif-filtering-guide/
# - https://cran.r-project.org/web/packages/CoordinateCleaner/vignettes/Cleaning_GBIF_data_with_CoordinateCleaner.html
# - https://www.r-bloggers.com/2021/03/downloading-and-cleaning-gbif-data-with-r/
if (!file_exists(derived_data("flora_occ.tsv"))) {
  message("Reacquiring GBIF occurrences...")
  flora <- mutate(
    flora,
    occ = map(
      taxon, gbif_data, 
      hasCoordinate = TRUE, hasGeospatialIssue = FALSE, geometry = w_asia, 
      limit = 100000, .progress = TRUE
    ),
    occ = map(occ, gbif_drop_imprecise_coords, threshold = 5000),
    occ = map(occ, gbif_drop_fossils),
    occ = map(occ, gbif_drop_duplicate_coords),
    occ = map(occ, gbif_standardise_names),
    occ = map(occ, \(x) {
      select(x, gbif_key = key, scientific_name, genus, species, longitude, 
             latitude)
    })
  )
  write_tsv(unnest(flora, cols = c(occ)), derived_data("flora_occ.tsv"))
}
flora <- nest(read_tsv(derived_data("flora_occ.tsv")), occ = gbif_key:latitude)
```

```{r fig-occ-map}
# TODO: Can this be made presentable?
#flora |>
#  unnest(occ) |>
#  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
#  ggplot() + 
#  facet_wrap(vars(taxon)) +
#  layer_spatial(ne_land, fill = "#ffffff") +
#  geom_sf(shape = ".") +
#  coord_sf(xlim = w_asia[c("xmin", "xmax")], 
#           ylim = w_asia[c("ymin", "ymax")],
#           crs = w_asia_albers,
#           default_crs = latlong)
```

```{r data-background}
# TODO: slow - cache
flora <- flora |>
  mutate(
    occ = map(occ, mutate, present = TRUE, weight = 1),
    bg = map(occ, function(x, region, n) { 
      background_sample(
        region = region, 
        N = n,
        coord_x = "longitude", coord_y = "latitude",
        present = FALSE,
        weight = (1 / n) * nrow(x) # weight presences and background equally
      )
    }, region = w_asia_land, n = n_background),
    # occ_bg = map2(occ, bg, bind_rows)
    occ = map2(occ, bg, bind_rows)
  ) |>
  select(-bg)
```

<!-- TODO: rewrite -->
Occurrence data only tells us where a species is present;
there is rarely definitive information on where the species is *not* found.
We therefore need to generate random background points or "pseudo-absences" to feed to the model.
There are several ways to do this. 
We follow the advice of @BarbetMassinEtAl2012 for regression-based species distribution models and use a large (:10000) random sample of points, weighted equally against the presences in the regression.
@ValaviEtAl2022 also recommend using a very large background sample for random forest models.

@@

climate <- read_stars(
  derived_data(
    "paleoclim", 
    paste0("paleoclim_", climate_periods$code, "_2_5m_w_asia.tif")
  ),
  along = NA
)

names(climate) <- climate_periods$code

@@

terrain <- read_stars(derived_data("dem", c("srtm_15plus_w_asia.tif", 
                                            "srtm_15plus_w_asia_slope.tif",
                                            "srtm_15plus_w_asia_aspect.tif",
                                            "srtm_15plus_w_asia_wetness.tif")))
names(terrain) <- c("elevation", "slope", "aspect", "wetness")
```

@@

```{r data-environment}
# Prepare environmental data cube for prediction (not training)
# Duplicate time-invariant predictors to create dummy period dimension
terrain_cube <- list(terrain) |>
  rep(nrow(climate_periods)) |>
  set_names(climate_periods$code) |>
  c(list(along = "period")) |>
  do.call(what = c)

soil_cube <- list(soil) |>
  rep(nrow(climate_periods)) |>
  set_names(climate_periods$code) |>
  c(list(along = "period")) |>
  do.call(what = c)

# Resample predictor data to common projection and resolution
climate_cube <- st_warp(climate, crs = w_asia_albers, 
                        cellsize = c(pred_res, pred_res))
terrain_cube <- st_warp(terrain_cube, climate_cube)
soil_cube <- st_warp(soil_cube, climate_cube)

# Combine into prediction cube
environment <- c(climate_cube, terrain_cube, soil_cube)

# Save some memory
rm(climate_cube, terrain_cube, soil_cube)
```

@@

```{r extract-predictor-data}
# TODO: factor out
bind_extract <- function(x, y) {
  extract <- st_extract(y, st_transform(x, st_crs(y)))
  cbind(x, select(tibble(extract), -geometry))
}

flora <- flora |> 
  mutate(
    occ = map(occ, st_as_sf, coords = c("longitude", "latitude"), crs = latlong, remove = FALSE),
    occ = map(occ, bind_extract, slice(climate, "period", "cur")),
    occ = map(occ, bind_extract, terrain),
    occ = map(occ, bind_extract, soil)
  )
```

@@

```{r niche-data}
# Split training, test (for model assessment) and cross-validation sets (for
# tuning)
flora <- flora |>
  mutate(
    # De-sf
    occ = map(occ, \(x) select(tibble(x), -geometry)),
    # TODO: can this happen earlier?
    occ = map(occ, \(x) mutate(x, present = factor(present, c(FALSE, TRUE)))),
    occ = map(occ, \(x) mutate(x, weight = importance_weights(weight))),
    # occ = map(occ, \(x) drop_na(x, starts_with("bio"))), # TODO: where do these come from?
    occ_split = map(occ, \(x) initial_split(x, strata = present)),
    # occ_folds = map(occ_split, \(x) vfold_cv(training(x), v = 10, strata = present))
  )

# Data preprocessing recipe
# Following recommended preprocessing steps for RF classification from:
#   https://www.tmwr.org/pre-proc-table.html
niche_recipe <- recipe(training(flora$occ_split[[1]])) |>
  update_role(gbif_key:latitude, new_role = "ID") |>
  update_role(present, new_role = "outcome") |>
  update_role(starts_with("bio"), new_role = "predictor") |>
  update_role(c(elevation, slope, aspect, wetness, clay, phh2o, sand, silt),
              new_role = "predictor") |>
  # Impute missing values based on spatial autocorrelation
  step_impute_linear(all_numeric_predictors(), 
                     impute_with = imp_vars(latitude, longitude)) |>
  # Drop redundant predictors
  step_nzv(all_predictors())
```

@@

niche_workflow <- workflow(niche_recipe, niche_model)
```

@@
